Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression

### 2022-06-27 docs/rmd/presentation/slides.Rmd
autogit

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
header-includes:
   - \usepackage{float}
---

@@

@@

* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behaviour of genetic programming in symbolic regression if compared to tournament selection?

@@

* A metaheuristic that searches for computer programs that solve a given problem^[@koza_main]
* Evolutionary Algorithm: Simulates the process of Darwinian evolution
* Unique Features:
  * Evolve solutions of variable length and structure
  * Solutions are typically represented by recursive tree structures

@@

### Torunament Selection

@@

* Objective: Create Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
* Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
* Intuition: Higher chance for "specialist" solutions to be selected since it is decided on a per case basis

@@

* Task: Find a mathematical model that fits a given set of datapoints
* One of the first applications of GP described by @koza_main
* High relevance: GP can outperform state-of-the-art machine learning algorithms like gradient boosting ^[@Orzechowski_2018]

@@

* Main objective in most supervised machine learning problems: Achieve good performance for unseen data
* Challenge: Avoid overfitting to training data
* Little attention has been paid to generalization in GP ^[@open_issues_gp, @generalisation_in_gp]

@@

```{r vars_dataset echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

@@

* Total dataset ($N=768$) is randomly split into a training and testing dataset (50:50)
* Fitness metric: Mean squared Error (MSE)
* Train two models using GP with the training dataset only, one using tournament selection and the other $\epsilon$-lexicase selection
* For each generation: Select elite model and compute its fitness for the testing dataset

@@

## Example

Model evolved by tournament selection after 100 generations:

![](../plots/example_model.png)

@@

```{r pvals_fitness, echo=FALSE, warning=FALSE}
df <- read.csv("../tables/csv/mwu_matrix_error.csv")

@@

* GP typical growth behaviour for both operators
* Solutions grow at a similiar rate in each generation
* No statistically significant differences in overall program size based on selection

@@

```{r ea_config, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

>>>>

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    theme: "Warsaw"
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
includes:
      in_header: header.tex
---

@@

\setbeamertemplate{footline}[frame number]

@@

* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behavior of genetic programming in symbolic regression if compared to tournament selection?

@@

* A meta-heuristic that searches for computer programs that solve a given problem^[@koza_main]
* Evolutionary Algorithm: Simulates the process of Darwinian evolution
* Unique Features:
  * Evolve solutions of variable length and structure
  * Solutions are typically represented by recursive tree structures

@@

### Tournament Selection

@@

* Objective: Create Selection method for uncompromising, continuous-valued symbolic regression problems ^[@6920034, p.12]
* Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
* Intuition: Higher chance for "specialist" solutions to be selected since it is decided on a per case basis

@@

* Task: Find a mathematical model that fits a given set of datapoints
* One of the first applications of GP described by @koza_main
* High relevance: GP can outperform state-of-the-art machine learning algorithms, e.g. gradient boosting ^[@Orzechowski_2018]

@@

* Main objective in most supervised machine learning problems: Achieve good performance for unseen data
* Challenge: Avoid overfitting to training dataset
* Little attention has been paid to generalization in GP ^[@open_issues_gp, @generalisation_in_gp]

@@

```{r vars_dataset, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

@@

* Total dataset ($N=768$) is randomly split into a training/testing dataset (50/50)
* Fitness metric: Mean squared Error (MSE)
* Train two models using GP with the training dataset only, one using tournament selection and the other $\epsilon$-lexicase selection
* For each generation: Select elite model and compute its fitness for the testing dataset

@@

@@

```{r pvals_fitness_table, echo=FALSE, warning=FALSE}
df <- read.csv("../tables/csv/mwu_matrix_error.csv")

@@

* GP typical growth behavior for both operators
* Solutions grow at a similar rate in each generation
* No statistically significant differences in overall program size based on selection

@@

```{r ea_config_table, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

### 2022-06-27 docs/rmd/presentation/slides.tex
autogit

\usepackage{float}

@@

\begin{frame}[allowframebreaks]
  \tableofcontents[hideallsubsections]
\end{frame}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\begin{frame}{Research Question}
\protect\hypertarget{research-question}{}
\begin{itemize}
\tightlist
\item
  Does the usage of \(\epsilon\)-lexicase parent selection influence the
  generalization behaviour of genetic programming in symbolic regression
  if compared to tournament selection?
\end{itemize}
\end{frame}

\begin{frame}{Genetic Programming}
\protect\hypertarget{genetic-programming}{}
\begin{itemize}
\tightlist
\item
  A metaheuristic that searches for computer programs that solve a given
  problem\footnote<.->{Koza (1992)}
\item
  Evolutionary Algorithm: Simulates the process of Darwinian evolution
\item
  Unique Features:

@@

\begin{frame}{Parent Selection}
\protect\hypertarget{parent-selection}{}
\begin{block}{Torunament Selection}
\protect\hypertarget{torunament-selection}{}
\begin{itemize}
\tightlist
\item
  Most commonly used selection operator in Genetic Programming
  (GP)\footnote<.->{Fang and Li (2010), p.181}
\item
  Intuition: High chance for ``generalist'' solutions to be selected
  since it is based on aggregated fitness scores
\end{itemize}
\end{block}

\begin{block}{\(\epsilon\)-Lexicase Selection}
\protect\hypertarget{epsilon-lexicase-selection}{}
\begin{itemize}
\tightlist
\item
  Objective: Create Selection method for uncompromising,
  continous-valued symbolic regression problems \footnote<.->{Helmuth,
    Spector and Matheson (2015), p.12}
\item
  Performance increases have been demonstrated in many benchmarking
  problems \footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
\item
  Intuition: Higher chance for ``specialist'' solutions to be selected
  since it is decided on a per case basis
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Example}
\protect\hypertarget{example}{}
Model evolved by tournament selection after 100 generations:

\includegraphics{../plots/example_model.png}
\end{frame}

@@

\caption{\label{tab:pvals_fitness}Mean Error - P-Values (MWU)}
\centering
\fontsize{4}{6}\selectfont
\begin{tabular}[t]{l|r|r|r|r}
\hline
X & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
\hline
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
\hline
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
\hline
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
\hline
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\begin{frame}{Finding 3}
\protect\hypertarget{finding-3}{}
\begin{itemize}
\tightlist
\item
  GP typical growth behaviour for both operators
\item
  Solutions grow at a similiar rate in each generation
\item
  No statistically significant differences in overall program size based
  on selection
\end{itemize}
\end{frame}

@@

\caption{\label{tab:ea_config}Evolutionary Parameters}
\centering
\begin{tabular}[t]{l|l}
\hline
Parameter & Value\\
\hline
Population Size & 500\\
\hline
Number of Generations & 100\\
\hline
Mutation Rate & 20\%\\
\hline
Crossover Rate & 80\%\\
\hline
Tournament Size & 3\\
\hline
Epsilon selection & automatic\\
\hline
Elite Size & 0\\
\hline
\end{tabular}
\end{table}
\end{frame}

>>>>

@@

\begin{frame}[allowframebreaks]
  \tableofcontents[hideallsubsections]
\end{frame}
\begin{frame}
\setbeamertemplate{footline}{[}frame number{]}
\end{frame}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\begin{frame}{Research Question}
\protect\hypertarget{research-question}{}
\begin{itemize}
\tightlist
\item
  Does the usage of \(\epsilon\)-lexicase parent selection influence the
  generalization behavior of genetic programming in symbolic regression
  if compared to tournament selection?
\end{itemize}
\end{frame}

\begin{frame}{Genetic Programming}
\protect\hypertarget{genetic-programming}{}
\begin{itemize}
\tightlist
\item
  A meta-heuristic that searches for computer programs that solve a
  given problem\footnote<.->{Koza (1992)}
\item
  Evolutionary Algorithm: Simulates the process of Darwinian evolution
\item
  Unique Features:

@@

\begin{frame}{Parent Selection}
\protect\hypertarget{parent-selection}{}
\begin{block}{Tournament Selection}
\protect\hypertarget{tournament-selection}{}
\begin{itemize}
\tightlist
\item
  Most commonly used selection operator in Genetic Programming
  (GP)\footnote<.->{Fang and Li (2010), p.181}
\item
  Intuition: High chance for ``generalist'' solutions to be selected
  since it is based on aggregated fitness scores
\end{itemize}
\end{block}

\begin{block}{\(\epsilon\)-Lexicase Selection}
\protect\hypertarget{epsilon-lexicase-selection}{}
\begin{itemize}
\tightlist
\item
  Objective: Create Selection method for uncompromising,
  continuous-valued symbolic regression problems \footnote<.->{Helmuth,
    Spector and Matheson (2015), p.12}
\item
  Performance increases have been demonstrated in many benchmarking
  problems \footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
\item
  Intuition: Higher chance for ``specialist'' solutions to be selected
  since it is decided on a per case basis
\end{itemize}
\end{block}
\end{frame}

@@

@@

\caption{\label{tab:pvals_fitness_table}Mean Error - P-Values (MWU)}
\centering
\fontsize{4}{6}\selectfont
\begin{tabular}[t]{l|r|r|r|r}
\hline
  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
\hline
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
\hline
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
\hline
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
\hline
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\begin{frame}{Finding 3}
\protect\hypertarget{finding-3}{}
\begin{itemize}
\tightlist
\item
  GP typical growth behavior for both operators
\item
  Solutions grow at a similar rate in each generation
\item
  No statistically significant differences in overall program size based
  on selection
\end{itemize}
\end{frame}

@@

\caption{\label{tab:ea_config_table}Evolutionary Parameters}
\centering
\begin{tabular}[t]{l|l}
\hline
Parameter & Value\\
\hline
Population Size & 500\\
\hline
Number of Generations & 100\\
\hline
Mutation Rate & 20\%\\
\hline
Crossover Rate & 80\%\\
\hline
Tournament Size & 3\\
\hline
Epsilon selection & automatic\\
\hline
Elite Size & 0\\
\hline
\end{tabular}
\end{table}
\end{frame}
